Reduced GW invariants from cuspidal curves

### 2017-12-04 draft.tex
second approach to the contraction

\newcommand{\M}[4]{\overline{\mathcal M}_{#1,#2}(#3,#4)}
\newcommand{\PP}{\mathbb P}
\newcommand{\OO}{\mathcal O}
\renewcommand{\to}{\rightarrow}
\newcommand{\pr}{\rm{pr}}
\newcommand{\A}{\mathcal A}
\newcommand{\B}{\mathcal B}
\newcommand{\X}{\mathcal X}
\newcommand{\XP}{\mathcal{XP}}
\renewcommand{\L}{\mathcal L}
\newcommand{\hL}{\hat{\mathcal L}}
\newcommand{\C}{\mathfrak C}
\newcommand{\cC}{\mathcal C}
\newcommand{\Z}{\mathcal Z}
\newcommand{\MM}{\mathfrak M}
\newcommand{\oM}{\overline{\mathcal M}}
\newcommand{\R}{\operatorname{R}}

@@

\end{enumerate}
If instead $\mathcal C_0$ has got an elliptic tail, then we have two possibilities:
\begin{enumerate}
\setcounter{enumi}{1}
\item $\mathcal C_{\eta}$ has an elliptic tail as well; that is the image of $S\to\MM$ is contained in the boundary, so we can find a lift
\bcd
 & \MM_{1,1}\times\MM_{0,1+n}^\text{wt}\ar[d] & \\
 S\ar[r]\ar[ur,dashed] & \mathfrak D_{\{1,\emptyset\},\{0,n\}}\ar[r,hook] &\MM.
\ecd
 Then $\mathcal C_S$ is the push-out of a family of rational curves $\mathcal R_S$ and a family of genus one curves $\mathcal Z_S$:
%The map $S\to\partial \mathfrak M_{1,n}$ lifts to the normalisation of the boundary, i.e. $\mathfrak M_{0,n_0+1}\times\mathfrak M_{1,n_1+1}$, which gives two families of pointed curves on $S$; there is a map from the push-out to $\mathcal C_S$ by the universal property, and it is proper bijective and flat (crit\`{e}re de platitude par fibres \cite[\href{http://stacks.math.columbia.edu/tag/039A}{Tag 039A}]{stacks-project}).
\begin{equation}\label{eq:pushout}
\begin{tikzcd}
S\ar[d]\ar[r] & \mathcal R_S \ar[d] \\
\mathcal Z_S\ar[r] & \mathcal C_S
\end{tikzcd}
\end{equation}
Recall that the cuspidal curve $\cC'_K$ can be characterised as the push-out of the following diagram:
\bcd
2K\ar[r]\ar[d] & \mathcal R_K\ar[d] \\
K\ar[r] & \mathcal C'_K.
\ecd
Since the smooth section $S\to\mathcal R_S$ defines a Cartier divisor, it makes sense to take its double and we can thus define $\cC'_S$ by means of the similar diagram:
\bcd
2S\ar[r]\ar[d] & \mathcal R_S\ar[d] \\
S\ar[r] & \mathcal C'_S
\ecd
The morphism $\phi_S\colon\cC_S\to\cC'_S$ extending $\phi_\eta$ is then defined by exploiting the description of $\cC_S$ as a push-out \eqref{eq:pushout}, and the morphisms $\id\colon\mathcal R_S\to \mathcal R_S$ and $\pr_{\mathcal Z_S}\colon\mathcal Z_S\to S$.
\item If $\mathcal C_S$ smoothens the elliptic tail, then $\phi_\eta$ is an isomorphism. We may assume that $S$ is the spectrum of a complete DVR with algebraically closed residue field \cite[Theorem 7.10]{LMB}. Then we may pick one smooth section for each rational component of $\cC_0$ and extend them to sections of $\cC_S\to S$ by Grothendieck's existence theorem; let us denote by $\Sigma$ the Cartier divisor that is the sum of all such sections. Let $Z$ be the elliptic tail in the central fiber; then we claim that $\omega_{\cC_S/S}(Z)\otimes\mathcal O_{\cC_S}(\Sigma)$ is $\pi_S$ semi-ample, ample on the generic fiber, and gives the contraction of the elliptic tail to the cusp in the central fiber. We shall not prove the claim here, since this is the core of the second approach.
\end{enumerate}
Finally, observe that the map is bijective by construction and $\mathfrak M$ is normal, hence $\pi_{|\mathcal X}\colon\mathcal X\to \mathfrak M$ is an isomorphism by Zariski's main theorem (as in \cite[\href{http://stacks.math.columbia.edu/tag/082I}{Tag 082I}]{stacks-project}).
 \end{proof}
\subsection*{Why only cusps?} 
\subsection*{Alternative constructions}
\begin{itemize}
\item adapt the ideas of \cite{HassettHyeon}: check out section 3 of this paper;
\item check \cite[\S 4]{RSPW}.
\end{itemize}
 \section{Comparison with $p$-fields and splitting}
\subsection{Notation}
Let $\Z$ be defined by the pullback diagram:
\bcd
\mathcal Z \ar[r]\ar[d]\ar[rd,phantom,"\Box"] & \overline{\mathcal M}_1^{(1)}(\PP^4,d) \ar[d] \\
\mathcal X\ar[r] & \mathfrak M_1^{(1)}
\ecd

>>>>

\newcommand{\M}[4]{\overline{\mathcal M}_{#1,#2}(#3,#4)}
\newcommand{\PP}{\mathbb P}
\newcommand{\OO}{\mathcal O}
\renewcommand{\to}{\rightarrow}
\newcommand{\pr}{\rm{pr}}
\newcommand{\Aaff}{\mathbb A}
\newcommand{\A}{\mathcal A}
\newcommand{\B}{\mathcal B}
\newcommand{\X}{\mathcal X}
\newcommand{\XP}{\mathcal{XP}}
\renewcommand{\L}{\mathcal L}
\newcommand{\hL}{\hat{\mathcal L}}
\newcommand{\C}{\mathfrak C}
\newcommand{\cC}{\mathcal C}
\newcommand{\Z}{\mathcal Z}
\newcommand{\MM}{\mathfrak M}
\newcommand{\pP}{\mathfrak P}
\newcommand{\oM}{\overline{\mathcal M}}
\newcommand{\R}{\operatorname{R}}
\newcommand{\Gm}{\mathbb{G}_{\rm{m}}}

@@

\end{enumerate}
If instead $\mathcal C_0$ has got an elliptic tail, then we have two possibilities:
\begin{enumerate}
\setcounter{enumi}{1}
\item $\mathcal C_{\eta}$ has an elliptic tail as well; that is the image of $S\to\MM$ is contained in the boundary, so we can find a lift
\bcd
 & \MM_{1,1}\times\MM_{0,1+n}^\text{wt}\ar[d] & \\
 S\ar[r]\ar[ur,dashed] & \mathfrak D_{\{1,\emptyset\},\{0,n\}}\ar[r,hook] &\MM.
\ecd
 Then $\mathcal C_S$ is the push-out of a family of rational curves $\mathcal R_S$ and a family of genus one curves $\mathcal Z_S$:
%The map $S\to\partial \mathfrak M_{1,n}$ lifts to the normalisation of the boundary, i.e. $\mathfrak M_{0,n_0+1}\times\mathfrak M_{1,n_1+1}$, which gives two families of pointed curves on $S$; there is a map from the push-out to $\mathcal C_S$ by the universal property, and it is proper bijective and flat (crit\`{e}re de platitude par fibres \cite[\href{http://stacks.math.columbia.edu/tag/039A}{Tag 039A}]{stacks-project}).
\begin{equation}\label{eq:pushout}
\begin{tikzcd}
S\ar[d]\ar[r] & \mathcal R_S \ar[d] \\
\mathcal Z_S\ar[r] & \mathcal C_S
\end{tikzcd}
\end{equation}
Recall that the cuspidal curve $\cC'_K$ can be characterised as the push-out of the following diagram:
\bcd
2K\ar[r]\ar[d] & \mathcal R_K\ar[d] \\
K\ar[r] & \mathcal C'_K.
\ecd
Since the smooth section $S\to\mathcal R_S$ defines a Cartier divisor, it makes sense to take its double and we can thus define $\cC'_S$ by means of the similar diagram:
\bcd
2S\ar[r]\ar[d] & \mathcal R_S\ar[d] \\
S\ar[r] & \mathcal C'_S
\ecd
The morphism $\phi_S\colon\cC_S\to\cC'_S$ extending $\phi_\eta$ is then defined by exploiting the description of $\cC_S$ as a push-out \eqref{eq:pushout}, and the morphisms $\id\colon\mathcal R_S\to \mathcal R_S$ and $\pr_{\mathcal Z_S}\colon\mathcal Z_S\to S$.
\item If $\mathcal C_S$ smoothens the elliptic tail, then $\phi_\eta$ is an isomorphism. We may assume that $S$ is the spectrum of a complete DVR with algebraically closed residue field \cite[Theorem 7.10]{LMB}. Then we may pick one smooth section for each rational component of $\cC_0$ and extend them to sections of $\cC_S\to S$ by Grothendieck's existence theorem; let us denote by $\Sigma$ the Cartier divisor that is the sum of all such sections. Let $Z$ be the elliptic tail in the central fiber; then we claim that $\omega_{\cC_S/S}(Z)\otimes\mathcal O_{\cC_S}(\Sigma)$ is $\pi_S$ semi-ample, ample on the generic fiber, and gives the contraction of the elliptic tail to the cusp in the central fiber. We shall not prove the claim here, since this is the core of the second approach.
\end{enumerate}
Finally, observe that the map is bijective by construction and $\mathfrak M$ is normal, hence $\pi_{|\mathcal X}\colon\mathcal X\to \mathfrak M$ is an isomorphism by Zariski's main theorem (as in \cite[\href{http://stacks.math.columbia.edu/tag/082I}{Tag 082I}]{stacks-project}).
 \end{proof}

\subsection*{Second approach: constructing the contraction}
The idea of this construction is essentially due to Hassett \cite[\S2]{HassettHyeon} and it has recently been reviewed and simplified in \cite[\S3.7]{RSPW}.

Let $\mathfrak{P}=\mathfrak{Pic}^d_{\cC/\MM_1}$ with the stability condition that $\omega_\pi\otimes \mathcal L^{\otimes 3}$ is $\pi$-ample. We work over $\MM_1^{\rm{div}}$, parametrising families of nodal curves with a relative Cartier divisor. More precisely, this can be thought of as the open inside $C(\pi_*\mathcal L)=\Spec_\pP(\R^1\pi_*\mathcal L)$ (see \cite{CLpfields}), where the section is not $0$ on any irreducible component of the curve. Alternatively one can think of the moduli functor of a line bundle with a section up to scalar as the hom-stack $\Hom_{\MM_1}(\C,[\Aaff^1/\Gm])$, again requiring the section not to be $0$ and the total degree of the line bundle to be $d$.

We shall first construct the contraction over $\MM^{\rm{div}}$ and then show that it descends to $\MM^{\rm{wt}}$.

Let $E$ be the locus inside the universal curve spanned by elliptic tails; this is a Cartier divisor in the universal curve over $\MM^{wt}_1$, which we shall freely pullback and keep denoting by $E$. We write $\mathcal M:=\omega_{\pi}(E)\otimes\mathcal O_{\cC}(3D)$, where $D$ is the universal Cartier divisor over $\MM_1^{\rm{div}}$ (so $\mathcal O_{\cC}(D)$ is the pullback of the universal line bundle on $\pP$).

\begin{prop}\label{1-stabilization-div}
 The $1$-stabilisation $\MM_1^{\rm{div}}\to\MM_1^{(1),\rm{div}}$ is defined by the line bundle $\mathcal M$:
 \bcd
 (\cC,D)\ar[rr,"\phi"]\ar[dr,"\pi"] & & (\hat\cC=\operatorname{Proj}_{\MM_1^{\rm{div}}}(\bigoplus_{n\geq 0}\pi_*\mathcal M^{\otimes n}),\phi(D))\ar[ld,"\hat\pi"] \\
 & \MM_1^{\rm{div}} &
 \ecd
\end{prop}
We prove a proposition through a series of Lemma s. The first one is a technical statement telling us that we can reduce to verify semi-ampleness and local freeness on families over DVR.
\begin{lem}{Reduction to DVR}\label{DVR}
Let $T$ the spectrum of a DVR with closed point $0$ and generic point $\eta$ and let $f\colon T\to\MM_1^{\rm{div}}$ such that $0\in \mathfrak{D}^1$ and $\eta\in \MM_1^{\rm{div}}\setminus \mathfrak{D}^1.$ Then 
\[f^*\pi_*\mathcal M\cong \pi_*f^*\mathcal M.\]
We have denoted with $\mathfrak{D}^1$ the Cartier divisor image of the locus spanned by elliptic tails.
\end{lem}
\begin{proof}
The argument is taken from \cite[Lemmma~3.7.2.2]{RSPW}.
Let $K^{\bullet}$ the complex of locally free sheaves on $S$ which satisfy cohomology and base change, i.e. such that for any $T\xrightarrow{f} S$  we get
$H^i(f^*K^{\bullet})=\R^i\pi_*f^*\mathcal M.$ The construction of $K^{\bullet}$ is standard, (see e.g. \cite[Proposition~12.2]{HAR} and since the cohomology is concentrated in degree 0,1, we can assume $K^{\bullet}=K_0\to K_1.$ Let $f$ be the map defined in the statement of the Lemma.
Then we have the following exact sequences:
\begin{equation}\label{eq:exact1}
0\to \pi_*\mathcal M\to K^0\to K^1\to \R^1\pi_*\mathcal M\to 0,
\end{equation}
\begin{equation}\label{eq:exact2}
0\to \pi_*f^*\mathcal M\to f^*K^0\to f^*K^1\to \R^1\pi_*f^*\mathcal M\to 0.
\end{equation}
The derived pull-back $Lf^*$ of the complex~\eqref{eq:exact1} is zero since the complex is exact. Writing down explicitly the spectral sequence which compute the derived pull-back, we find that its convergence to zero implies that:
\begin{align*}
\operatorname{coker}(f^*\pi_*\mathcal M\to\pi_*f^*\mathcal M)&\cong \rm{L}_1f^*\R^1\pi_*\mathcal M\\
\operatorname{ker}(f^*\pi_*\mathcal M\to\pi_*f^*\mathcal M)&\cong \rm{L}_2f^*\R^1\pi_*\mathcal M.
\end{align*} 
Now, notice that $\R^1\pi_*\mathcal M$ is a line bundle supported on the Cartier divisor $\mathfrak{D}^1;$ it is clearly zero on the smooth locus and one can easily check that is rank is constant and equal to one on the divisor.
To prove that $\rm{L}_if^*\R^1\pi_*\mathcal M=0$ for $i=1,2$ it is enough to work locally on $S$, where we have a free resolution for  of the form $\R^1\pi_*\mathcal M$ of the form:
\[0\rightarrow \OO_S\xrightarrow{g}\OO_S\rightarrow \R^1\pi_*\mathcal M\rightarrow 0\] 
and thus pulling back along $f$
\[0\rightarrow  \rm{L}_1f^*\R^1\pi_*\mathcal M \OO_T\xrightarrow{f^*g}\OO_T\rightarrow f^*\R^1\pi_*\mathcal M\rightarrow 0.\] 
It immediately follows that $ \rm{L}_2f^*\R^1\pi_*\mathcal M=0.$ Finally notice that since the image of $f$ is not contained in the boundary divisor $f^*g\neq 0$ and thus injective of sheaves as $T$ is a DVR. We can then conclude $ \rm{L}_1f^*\R^1\pi_*\mathcal M=0$ as well which implies the Lemma.
\end{proof}
\begin{lem}\label{lemma:semiample}
The line bundle $\mathcal M$ is $\pi$ semi-ample, i.e. the natural map
\[\pi^*\pi_*\mathcal M^{\otimes n}\to \mathcal M^{\otimes n}\]
is surjective.
\end{lem}
\begin{proof}
Outside the locus of elliptic tails $\mathcal M$ is $\pi$- ample and thus in particular $\pi$-semi ample. To check that it is still $\pi$-semi ample also for any $x\in C$ when $C$ has an elliptic tail, thanks to the above Lemma we can reduce to check this condition when $C$ is the central fiber of a one parameter smoothing of the elliptic tail. But then this has been proved by Smyth~\cite[Lemma~2.12]{SMY1}.
\end{proof}

\begin{lem}
$\pi_*\mathcal M$ is locally free of rank $d$ on $\MM_1^{\rm{div}}.$
\end{lem}
\begin{proof}\cite[Proposition~3.7.2.1]{RSPW}
We have essentially to check that $\pi_*\mathcal M$ has constant rank.
On $\MM_1^{\rm{div}}\setminus \mathfrak{D}^1$ this follows simply from the fact that $\R^1\pi_*\mathcal M=0$ and the rank is determined by Riemann-Roch.
Given a point $x$ on the boundary $\mathfrak{D}^1$, we can always find a DVR $T$ whose closed point maps to $x$ and whose generic point maps to $\MM_1^{\rm{div}}\setminus \mathfrak{D}^1.$ Then we are in the hypothesis of Lemma~\ref{DVR} and thus we can check the rank looking at $\pi_*f^*\mathcal M$ over the DVR  $T.$ But notice that $\pi_*f^*\mathcal M$ is torsion free over $T$, which conclude the proof.
\end{proof}

\begin{proof}\ref{1-stabilization-div}
Let $S\to \MM_1^{\rm{div}}$ be a smooth atlas, then we have:
 \bcd
 (C,D)\ar[rr,"\phi"]\ar[dr,"\pi"] & & (\hat C=\operatorname{Proj}_{S}(\bigoplus_{n\geq 0}\pi_*\mathcal M^{\otimes n}),\phi(D))\ar[ld,"\hat\pi"] \\
 & S &
 \ecd
where: $\phi$ is proper and birational  since $\mathcal M$ is $\pi$-semi-ample and $\hat{\pi}$ is flat since $\pi_*\mathcal M$ is locally free.

Consider the caresian diagram 
\bcd
C'\ar[r,"f_C"]\ar[d,"\pi'"] & C\ar[d,"\pi"] \\
S'\ar[r,"f"] & S.
\ecd
To show that the above construction is compatible with base change it is enough to show that for any base change holds what we proved in Lemma~\ref{DVR}:
\begin{equation}\label{eq:generalbasechange}
f^*\pi_*\mathcal M\cong \pi'_*f^*\mathcal M.
\end{equation}
Since $\pi$ is flat we do have an isomorphism if we consider everything derived:
\begin{equation}\label{eq:derivedbasechange}
\rm{L}f^*\R\pi_*\mathcal M\cong \R\pi'_*\rm{L}f^*\mathcal M.
\end{equation}
First notice that the $H^0$ on the LHS of \eqref{eq:derivedbasechange} is exactly the LHS of \ref{eq:generalbasechange}.

\end{proof}

\begin{prop}
The $1$-stabilisation with divisors descends to the level of weighted curves:
\bcd
\MM_1^{\rm{div}}\ar[r]\ar[d] & \MM_1^{(1),\rm{div}}\ar[d] \\
\MM_1^{\rm{wt}}\ar[r] & \MM_1^{(1),\rm{wt}}
\ecd
\end{prop}

\subsection*{Why only cusps?} 
\subsection*{Alternative constructions}
\begin{itemize}
\item adapt the ideas of \cite{HassettHyeon}: check out section 3 of this paper;
\item check \cite[\S 4]{RSPW}.
\end{itemize}
 \section{Comparison with $p$-fields and splitting}
\subsection{Notation}
Let $\Z$ be defined by the pullback diagram:
\bcd
\mathcal Z \ar[r]\ar[d]\ar[rd,phantom,"\Box"] & \overline{\mathcal M}_1^{(1)}(\PP^4,d) \ar[d] \\
\mathcal X\ar[r] & \mathfrak M_1^{(1)}
\ecd
